Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence.

### 2022-12-08 reports/archived/manuscript_v3.Rmd
archiving old scripts, manuscript versions and figures

**To complete after written**

@@

The four most widely sampled pathogen species/families were were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.). Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species. Among species with more than 500 samples the highest rates of infection were among *A. niloticus* (16%), *M. huberti* (11%) and *M. erythroleucus* (8%). The five most widely assayed species were sampled from a relative small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%). Infection with *Lassa mammarenavirus* was assessed in 28 species, with evidence of infection in 7 rodent species. Among species with more than 100 samples the highest rates of infection were among *M. natalensis* (22%), *M. erythroleucus* (6%) and *R. rattus* (4%). Species sampling occurred from limited areas within IUCN ranges (2-3%), however, sampling was more representative for rodent trapping studies (3-24%). Infection with species of Leptospiraceae was assessed in 8 species, with evidence of infection in 5 rodent species. Among species with more than 100 samples *M. natalensis* (19%), *R. rattus* (11%) and *Crocidura olivieri* (10%) were most frequently infected. Species sampling occurred across less than 1% of these species IUCN ranges and for all except *R. norvegicus* for less than 1% of their detection range in included studies. *Toxoplasma gondii* infection was assessed in 8 species, with evidence of infection in 6 rodent species. Among species with more than 100 samples *M. musculus* (7%), *M. erythroleucus* (4%) and *M. natalensis* (2%) were most commonly infected. Sampling occurred across less than 6% of their IUCN ranges, however, for *M. musculus*, *R. rattus* and *M. erythroleucus* sampling occurred across greater than 12% of their detection range.

@@

table_2 <- read_rds(here("tables", "Table_2_updated.rds"))

@@

@@

We synthesised data from 126 rodent trapping studies containing information on more than 72,000 trapped rodents across 1,331 trapping sites producing an estimated 966,000 trap-nights. Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level 2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes. For several important rodent species we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R. rattus* and *M. musculus*). We compared host-pathogen associations from rodent trapping studies with a global dataset and identified missing associations for several important zoonotic infectious diseases including, Lassa fever, Toxoplasmosis and Q fever in commonly occurring rodent species. Finally, we assessed the spatial coverage of testing for important zoonotic pathogens and found low representativeness across their host species range.

Rodent trapping studies provide important information on the distribution of rodent species and their assemblages, we have identified important bias in the locations in which studies to data have been performed with large regions of countries including Nigeria, Guinea and Cote d'Ivoire being understudied. Trapping sites are clustered in areas with relatively high human population density and anthropogenic land use which can importantly bias our understanding of the distribution of rodent species and their pathogens across West Africa. Further effort is required to minimise this bias by sampling currently undersampled regions.

Global datasets of rodent host distribution (e.g., IUCN) and presence (e.g., GBIF) provide useful resources to model the hazard of zoonotic disease spillover and novel pathogen emergence. We show here that non-detection data from rodent trapping studies can provide additional information to describe the heterogeneity of rodent assemblages across diverse habitats in West Africa. Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodents range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.

Host-pathogen association datasets are becoming increasing comprehensive alongside increased access to molecular discovery techniques for potential pathogens. We show that currently available data from rodent trapping studies can supplement these resources to better understand the potential hosts and reservoirs of known zoonotic pathogens. Rodent trapping data can support studies using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.

We have identified important limitations in the sampling of rodent zoonoses for known diseases across West Africa. For example, studies of *Lassa mammarenavirus* have taken place across 3% of *M. natalensis*'s range in West Africa. Lassa fever is known to be endemic in eight West African countries, although it is potentially more widely spread. Further studies are required to investigate this pathogen across a wider geographic range. Other important zoonotic diseases including Leptospirosis and Toxoplasmosis have been assessed over very limited regions of their hosts distribution.

We propose that rodent trapping studies can provide important data to supplement global datasets to quantify the hazard of zoonotic disease spillover and novel pathogen emergence. They can reduce some of the bias in global datasets however, important limitations remain. Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a hosts range.

# Supplementary

## Supplementary figure 3

```{r comparing-host-pathogen-interactions-supplementary, out.width = "120%", fig.cap = "Figure 3. L) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified host-pathogen associations at pathogen family level through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset."}

knitr::include_graphics(here("figures", "Supplementary_figure_3_updated.png"))

@@

>>>>

Rodents are important important globally distributed reservoirs of known and novel zoonotic pathogens. Ongoing anthropogenic land use change is altering the composition of host species assemblages and increasing the risk of zoonotic disease spillover. These rapid changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk. However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are limited by inclusion of limited data sources **further limitations** and are often biased. Here, we synthesise data from West Africa from 128 rodent trapping studies, from 1964-2021, including over 72,000 individual rodent detections in order to more fully categorise the range and presence of important rodent reservoir species in this region. We identify that, although rodent trapping studies are biased towards human dominated landscapes across West Africa, we find that they can usefully complement current rodent species distribution datasets. For four regionally important zoonotic pathogens (*Borrelia spp.*, *Lassa mammarenavirus*, *Leptospira spp.* and *Toxoplasma gondii*) we identify host-pathogen associations that have not been reported in the consolidated CLOVER dataset. These omissions may bias studies investigating the effect of land use and climate change on future disease hazard. Finally, we find that for these four diseases the proportion of a rodent hosts range that has been sampled remains small. A priority of future rodent trapping studies should be to sample rodent hosts of these diseases across a greater geographic range to investigate the hazard of zoonotic spillover across West Africa. In the interim, future studies must incorporate a measure of the current sampling biases. The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support the production of more complete understanding of the hazard of zoonotic spillover events.

@@

The four most widely sampled pathogen species/families were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.). Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species. Among species with more than 500 samples the highest rates of infection were among *A. niloticus* (16%), *M. huberti* (11%) and *M. erythroleucus* (8%). The five most widely assayed species were sampled from a relatively small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%).

Infection with *Lassa mammarenavirus* was assessed in 28 species, with evidence of infection in 7 rodent species. Among species with more than 100 samples the highest rates of infection were among *M. natalensis* (22%), *M. erythroleucus* (6%) and *R. rattus* (4%). Species sampling occurred from limited areas within IUCN ranges (2-3%), however, sampling was more representative for rodent trapping studies (3-24%).

Infection with species of Leptospiraceae was assessed in 8 species, with evidence of infection in 5 rodent species. Among species with more than 100 samples *M. natalensis* (19%), *R. rattus* (11%) and *Crocidura olivieri* (10%) were most frequently infected. Species sampling occurred across less than 1% of these species IUCN ranges and for all except *R. norvegicus* for less than 1% of their detection range in rodnet trapping studies.

*Toxoplasma gondii* infection was assessed in 8 species, with evidence of infection in 6 rodent species. Among species with more than 100 samples *M. musculus* (7%), *M. erythroleucus* (4%) and *M. natalensis* (2%) were most commonly infected. Sampling occurred across less than 6% of their IUCN ranges, however, for *M. musculus*, *R. rattus* and *M. erythroleucus* sampling occurred across greater than 12% of their detection range.

<!---BLOCK_LANDSCAPE_START--->

@@

table_2 <- read_rds(here("tables", "Table_2_updated.rds"))%>%
  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 4 most widely sampled pathogens and the 5 most sampled potential rodent host species. * no IUCN range in West African") %>%
  autofit()

@@

<!---BLOCK_LANDSCAPE_STOP--->

@@

We synthesised data from 126 rodent trapping studies containing information on more than 72,000 trapped rodents across 1,331 trapping sites producing an estimated 966,000 trap-nights. Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level-2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes. For several important rodent species, we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R. rattus* and *M. musculus*). We compared host-pathogen associations from rodent trapping studies with a global dataset and identified missing associations for several important zoonotic infectious diseases including, Lassa fever, Toxoplasmosis and Q fever in commonly occurring rodent species. Finally, we assessed the spatial coverage of testing for important zoonotic pathogens and found low representativeness across their host species range.

Rodent trapping studies provide important information on the distribution of rodent species and their assemblages, we have identified important bias in the locations in which studies to data have been performed with large regions of countries including Nigeria, Guinea and Cote d'Ivoire being understudied. Trapping sites are clustered in areas with relatively high human population density and anthropogenic land use which can importantly bias our understanding of the distribution of rodent species and their pathogens across West Africa. Further effort is required to counteract this bias by sampling currently undersampled regions.

Global datasets of rodent host distribution (e.g., IUCN) and presence (e.g., GBIF) provide useful resources to model the hazard of zoonotic disease spillover and novel pathogen emergence. We show here that non-detection data from rodent trapping studies can provide additional information to describe the heterogeneity of rodent assemblages across diverse habitats in West Africa. Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodent's range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.

Host-pathogen association datasets are becoming increasing comprehensive alongside increased access to molecular discovery techniques for potential pathogens. We show that currently available data from rodent trapping studies can enrich these resources, particularly with prevalence data and geolocation to better understand the potential hosts and reservoirs of known zoonotic pathogens. Most of the host-pathogen associations from CLOVER not identified from rodent trapping studies occurred in species with ranges that go beyond our study area, suggesting that these pathogens do not exist in West Africa. Rodent trapping data can support analyses using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.

We have identified important limitations in the sampling of rodent zoonoses for known diseases across West Africa. For example, studies of *Lassa mammarenavirus* have taken place across 3% of *M. natalensis*'s range in West Africa. Lassa fever is known to be endemic in eight West African countries, although it is potentially more widely spread. Further rodent trapping studies are required to investigate this pathogen across a wider geographic range. Other important zoonotic diseases including Leptospirosis and Toxoplasmosis have been assessed over very limited regions of their hosts distribution.

Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies. Land use change and population density have changed dramatically over this period in West Africa. We attempted to mitigate this by using the median year of trapping to understand the spatial biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the study was conducted and the situation in 2005. Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to importantly change.

We propose that rodent trapping studies can provide important data to supplement global datasets to quantify the hazard of zoonotic disease spillover and novel pathogen emergence. They can reduce some of the bias in global datasets however, important limitations remain. Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a host's range.

# References

<div id="refs"></div>

# Supplementary
<!---BLOCK_LANDSCAPE_START--->
## Supplementary Table 1

```{r supplementary-table-1}

read_rds(here("tables", "supplementary_table_1.rds")) %>%
  set_caption(caption = "Supplementary Table 1: Data extraction tool for studies meeting inclusion criteria")

@@

## Supplementary Table 2

```{r supplementary-table-2}

read_rds(here("tables", "supplementary_table_2.rds"))

```
<!---BLOCK_LANDSCAPE_STOP--->
## Supplementary Table 3

```{r supplementary-table-3-1}

read_rds(here("tables", "supplementary_table_3_1.rds"))

```{r supplementary-table-3-2}

read_rds(here("tables", "supplementary_table_3_2.rds"))

```{r supplementary-table-3-3}

read_rds(here("tables", "supplementary_table_3_3.rds"))

```{r supplementary-table-3-4}

read_rds(here("tables", "supplementary_table_3_4.rds"))

## Supplementary Figure 1

![Supplementary Figure 1. Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity. Red points indicate the publication of studies. Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_figure_1.png")`)

## Supplementary Figure 2

![Supplementary Figure 2. Comparison between proportion of land use between regions trapped and all land use in West Africa. Cropland, Shrubland and Grassland were trapped at than representative rates while Bare habitats, Forested regions and Urban landscapes were trapped at lower than representative rates.](`r here("figures", "Supplementary_figure_2.png")`)

## Supplementary Figure 3

![Supplementary Figure 3. Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Supplementary_figure_3.png")`)

## Supplementary Figure 4

![Supplementary Figure 4. L) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified host-pathogen associations at pathogen family level through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supplementary_figure_4_updated.png")`)

### 2022-12-08 reports/archived/manuscript_v4.Rmd
archiving old scripts, manuscript versions and figures

Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (Han et al. 2015). Here we have synthesised data from 126 rodent trapping studies containing information on more than 76,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries. Locations studied are complementary to curated datasets (e.g. IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets (Boakes et al., 2010). We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset. However, pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots (Harvey and Holmes, 2022).

We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases (Beck et al., 2014). Relative trapping effort was biased towards increased effort in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion of non-native rodent species (M. musculus and R. rattus) and the risk of endemic zoonosis outbreaks (e.g. Lassa mammarenavirus). Much of West Africa remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges. For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana (Basinski et al., 2021). Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.

Rodent trapping studies provide geographic and temporally contextualised data on both species detection and non-detection which are not available from curated datasets. Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences (Václavík and Meentemeyer, 2009). Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk (Plowright et al., 2019).

Currently available consolidated datasets on host-pathogen associations (e.g. CLOVER, EID2 and GMPD2) do not include spatial or temporal components (Gibb, Gregory F. Albery, et al., 2021a). The current synthesis of rodent trapping studies has highlighted that pathogens have been sparsely sampled within a hosts range. Current zoonosis risk models dependent on these sources of data are therefore not able to incorporate spatial heterogeneity in pathogen prevalence across the host range. Additional uncertainty in current models of zoonotic disease risk arises from observed host-pathogen associations that have not been reported in these consolidated datasets. For example, Hylomyscus pamfi infected with Lassa mammarenavirus and R. rattus infected with Coxiella burnetiid, will not be included when solely dependent on consolidated host-pathogen datasets. Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses (Wilkinson et al., 2019).

Few studies stratified detection and non-detection of hosts or pathogen prevalence by time, therefore limiting inference of changes in host and pathogen dynamics. This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space (Fichet-Calvet et al., 2016). Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.

Finally, due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies. Land use change and population density have changed dramatically over this period in West Africa (Herrmann et al., 2020). We attempted to mitigate against this by using the median year of trapping as a reference point to understand the spatial and land use biases in trapping activity. It is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the land use and human population variables in 2005. Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change (Beck et al., 2014).

We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data. Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence. Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication (Foster and Deardorff, 2017).

Future rodent trapping studies should be targeted towards regions that are currently under-studied. Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence. Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis. Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events. Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement curated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.

>>>>

Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015). Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights. Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g. IUCN, GBIF). The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@boakes_distorted_2010]. We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset. Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@harvey_diversity_2022].

We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@beck_spatial_2014]. Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M. musculus* and *R. rattus*) and the hazard of endemic zoonosis outbreaks (e.g. Lassa mammarenavirus). Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges. For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@basinski_bridging_2021]. Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.

Rodent trapping studies provide data on both presence and absence which are not available from curated datasets. Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@vaclavik_invasive_2009]. Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.

Currently available consolidated datasets on host-pathogen associations (e.g. CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@gibb_data_2021]. The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range. Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range. Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R. rattus* and Q fever (*Coxiella burnetii*). Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@wilkinson_comparison_2019].

Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics. This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet-calvet_spatial_2016]. Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.

Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies. Land use change and population density have changed dramatically over this period in West Africa [@herrmann_accelerating_2020]. We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005. Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change.

We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data. Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence. However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@foster_open_2017].

Future rodent trapping studies should be targeted towards regions that are currently under-studied. Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence. Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis. Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
